A novel method for detecting coalescing binaries in near realtime with Advanced LIGO and beyond \editorial{Submit to PRD or CQG?}

### 2011-02-21 analysis.tex
Implementing my changes
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@658 ccdd05d5-56c1-461a-bf98-85f6876523b4

\section{Implementation and Analysis}
\label{SECIII}\label{sec:analysis}

\editorial{Should I mention GStreamer at all?} The detection method described above is much more efficient in terms of floating point operations than the traditional matched filter bank method.  However, time slices and conditional reconstruction greatly complicate queueing, synchronizing, and bookkeeping of intermediate signals.  A low latency implementation capable of recruiting more than one \textsc{cpu} core would be difficult to achieve within the familiar serial programming framework because of the nontrivial time-delay relationships between samples.  Due to these complications, we chose to prototype the search using an open source signal processing environment called GStreamer \cite{gstreamer}.  Primarily used for playing, authoring, or streaming media on Linux systems, GStreamer is an integral component of the popular Gnome desktop.

\editorial{It would be good to illustrate the layout of this particular template bank: masses spanned, time slice layout \dots}
\editorial{Drew: I think this should say the FFT version of LLOID is almost 10 times faster than the conventional FFT method.}
\begin{table}[!h]
\caption{Operation counts per sample for six different detection methods.  The operation counts for \textsc{lloid} assume a reconstruction duty cycle of 5\%.  Note that the \textsc{fir} method with \textsc{lloid} is almost 10 times faster than the conventional \textsc{fft} method, despite having substantially lower latency.}
\begin{center}
\input{flop_budget}
\end{center}
\end{table}

>>>>

\section{Software Framework}
\label{SECIII}\label{sec:analysis}

The detection method described above is much more efficient in terms of
floating point operations than the traditional matched filter bank method.
However, time slices and conditional reconstruction greatly complicate
queueing, synchronizing, and bookkeeping of intermediate signals.  A low
latency implementation capable of recruiting more than one \textsc{cpu} core
would be difficult to achieve within the familiar serial programming framework
because of the nontrivial time-delay relationships between samples.  Due to
these complications, we chose to prototype the search using an open source
signal processing environment called GStreamer \cite{gstreamer}.  Primarily
used for playing, authoring, or streaming media on Linux systems, GStreamer is
an integral component of the popular Gnome desktop.

### 2011-02-21 appendix.tex
Adding 0's in from of decimal places for readability
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@656 ccdd05d5-56c1-461a-bf98-85f6876523b4
Adding comment on aLIGO noise
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@657 ccdd05d5-56c1-461a-bf98-85f6876523b4
Implementing my changes
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@658 ccdd05d5-56c1-461a-bf98-85f6876523b4

The filter bank described below reproduces the ``zero detuning, high power'' Advanced \textsc{ligo} noise model of  \cite{Shoemaker:2009p9770} very faithfully.  Since it is composed of a small number of third order or lower linear filters, a digital implementation of it can produce mock data in realtime with negligibly few floating point operations.

\begin{eqnarray}
\label{eq:iirbank}
\fl x_i &[n] \sim& \mathcal{N}[0, 1] \quad \forall \, i, n \nonumber\\\noalign{\vskip 2mm}
\fl y_1 &[n] =& \left(4 \times 10^{-28}\right) x_1[n] - y_1[n-1] + 2 \cdot .99995 \cos \left(\frac{2\pi \cdot 9.103}{16384}\right) y_1[n-2] - .99995^2 y_1[n-3] \nonumber\\\noalign{\vskip 2mm}
\fl y_2 &[n] =& \left(1.1 \times 10^{-23}\right) (x_2[n] - x_2[n-1]) \nonumber\\\noalign{\vskip 2mm}
\fl y_3 &[n] =& \left(10^{-27}\right) x_3[n] - y_3[n-1] + 2 \cdot .999 y_3[n-2] - .999^2 y_3[n-3] \nonumber\\\noalign{\vskip 2mm}
\fl y_4 &[n] =& \left(4 \times 10^{-26}\right) x_4[n] - y_4[n-1] + 2 \cdot .87 \cos \left(\frac{2\pi \cdot 50}{16384}\right) y_4[n-2] - .87 ^ 2 y_4[n-3] \nonumber\\\noalign{\vskip 2mm}
\fl y_5 &[n] =& \left(6.5 \times 10^{-24}\right) x_5[n] - y_5[n-1] - 2 \cdot .45 y_5[n-2] - .45 ^ 2 y_5[n-3] \nonumber\\\noalign{\vskip 2mm}
\fl \; y &[n] =& \frac{3\sqrt{16384}}{4} \left(y_1[n] + y_2[n] + y_3[n] + y_4[n] + y_5[n]\right)
\end{eqnarray}

\section{Floating point operation counts}

@@

@@

>>>>

\section{Floating point operation counts}
\label{appendix:cost}

@@

\begin{comment}

% I think that this part can be swalled into the computational costs discussion above.

\subsection{Comparison of computational costs and latency}
Convolutions are best done with the use of the Fast Fourier Transform (FFT).
However there is an inherent latency in performing convolutions this way. This
section describes the computational cost and latency of FFT convolutions, time 
domain (TD) convolutions and TD convolutions with SVD.

Convolving one filter with data using FFTs involves 
$\mathcal{O}[10 N_s \log_2(N_s)]$
floating point operations where $N_s$ is the number of sample points in
the data being convolved and results in $N_s - N_f$ samples of usable output.  
In comparison time domain convolutions require
$\mathcal{O}[N_s N_f]$ where $N_f$ is the number of sample points in the
filter.  The FFT is faster to compute when $N_f > 10 \log_2(N_s)$.  However
the FFT has an inherent latency of $\mathcal{O}[N_s/f_s]$ seconds, where
$f_s$ is the sample rate when minimally overlapping FFTs are computed. It is
possible to reduce the latency of the FFT by performing vastly overlapping
FFTs.  The cost of such a procedure would be 
$\mathcal{O}[10 N_o N_s \log_2(N_s)]$.  Thus for comparable computational
cost where $N_f \approx 10 N_o \log_2(N_s)$ the latency of the FFT method 
would be $\mathcal{O}[N_s/f_s/N_o]$ seconds.  Depending on the task at hand,
latency requirements and computational costs can dictate what procedure to
use.

@@

The filter bank described below reproduces the ``zero detuning, high power'' Advanced \textsc{ligo} noise model of  \cite{Shoemaker:2009p9770} very faithfully.  Since it is composed of a small number of third order or lower linear filters, a digital implementation of it can produce mock data in realtime with negligibly few floating point operations. \editorial{dk: We should add the seismic wall to this PSD. I know this is waiting on switching to gsteamer 0.10.32 but people will worry about this otherwise.}

\begin{eqnarray}
\label{eq:iirbank}
\fl x_i &[n] \sim& \mathcal{N}[0, 1] \quad \forall \, i, n \nonumber\\\noalign{\vskip 2mm}
\fl y_1 &[n] =& \left(4 \times 10^{-28}\right) x_1[n] - y_1[n-1] + 2 \cdot 0.99995 \cos \left(\frac{2\pi \cdot 9.103}{16384}\right) y_1[n-2] - 0.99995^2 y_1[n-3] \nonumber\\\noalign{\vskip 2mm}
\fl y_2 &[n] =& \left(1.1 \times 10^{-23}\right) (x_2[n] - x_2[n-1]) \nonumber\\\noalign{\vskip 2mm}
\fl y_3 &[n] =& \left(10^{-27}\right) x_3[n] - y_3[n-1] + 2 \cdot 0.999 y_3[n-2] - 0.999^2 y_3[n-3] \nonumber\\\noalign{\vskip 2mm}
\fl y_4 &[n] =& \left(4 \times 10^{-26}\right) x_4[n] - y_4[n-1] + 2 \cdot 0.87 \cos \left(\frac{2\pi \cdot 50}{16384}\right) y_4[n-2] - 0.87 ^ 2 y_4[n-3] \nonumber\\\noalign{\vskip 2mm}
\fl y_5 &[n] =& \left(6.5 \times 10^{-24}\right) x_5[n] - y_5[n-1] - 2 \cdot 0.45 y_5[n-2] - 0.45 ^ 2 y_5[n-3] \nonumber\\\noalign{\vskip 2mm}
\fl \; y &[n] =& \frac{3\sqrt{16384}}{4} \left(y_1[n] + y_2[n] + y_3[n] + y_4[n] + y_5[n]\right)
\end{eqnarray}

### 2011-02-21 method.tex
Implementing my changes
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@658 ccdd05d5-56c1-461a-bf98-85f6876523b4

In the rest of the section, we will describe a detection strategy that makes
use of some very general properties of inspiral template banks in order to
evaluate a matched filter bank with far lower computational cost than the
\textsc{fir} method and far less latency than the \textsc{fft} method.

@@

If the template is known to be quasi-bandlimited, then the partial matched
filter for each time slice may be processed at a reduce sample rate.  Compact
binary inspiral waveforms fit this description superbly: they are chirps whose
frequency and amplitude rise according to power laws of time.  They are not
only quasi-bandlimited, but quasi-monochromatic.  By exploiting this property,
we can implement a matched filter for an inspiral waveform using a multirate
filter bank that requires far fewer floating point operations.

@@

Our second innovation exploits the fact that the templates in inspiral template
banks are, by design, highly correlated.  It is possible to greatly reduce the
number of matched filters required to achieve a particular minimum match by
designing an appropriate set of orthonormal {\em basis templates}.  A purely
numerical technique based on the application of the singular value
decomposition (\textsc{svd}) to inspiral waveforms is demonstrated by the
authors in~\cite{Cannon:2010p10398}.

\editorial{I'm sweeping the complex nature of the templates under the rug here.
Is it OK to regard our packing of the real and imaginary parts of the
\textsc{svd} as an implementation detail?  It makes the math a little more
concise.}  One may regard a bank of $N$ discretely sampled templates of length
$M$ samples, $h^\mathrm W(\theta_i; t_j) = [\mathbf H]_{ij} = H_{ij}$, as a
matrix.  The singular value decomposition is an exact factorization that exists
for any matrix such that
%
\begin{equation} \label{eq:svd-exact}
%
H_{ij} = [\mathbf {V \Sigma U}]_{ij} = \sum_{k=1}^{N} v_{ik} \sigma_k u_{kj},
%
\end{equation}
%
where $\mathbf V$ and $\mathbf U$ are both unitary matrices and $\mathbf
\Sigma$ is a diagonal matrix.  For our purposes, we associate the rows of
$\mathbf U$ with a minimal set of basis templates, which become the kernels of
basis filters.  These filters give rise to the orthogonal \textsc{snr}s,
$[\mathbf \rho^\perp]_k = \rho_k^\perp = \mathbf{U}_k \star x^\mathsf{W}$.  The
rows of $\mathbf{V \Sigma}$ become reconstruction coefficients that map linear
combinations of the orthogonal \textsc{snr}s from the basis filters back onto
\textsc{snr}s for the original templates of interest: $\rho_i = \sum_k [\mathbf
V \mathbf \Sigma]_{ik} \rho_k^\perp$.

In many applications of the \textsc{svd}, including ours, the matrix can be
well approximated by truncating the summation in equation~(\ref{eq:svd-exact})
at $L \ll N$:
%
\begin{equation}
%
H_{ij}^\prime = \sum_{k=1}^{L} v_{ik} \sigma_k u_{kj}
%
\end{equation}
%
The cumulative sum of squares of the {\em singular values}, $\sigma_k$,
measures the Frobenius norm of the approximation, such that
%
\begin{equation}
%
\frac{\| \mathbf H^\prime \|}{\| \mathbf H \|} = \left(\sum_{k=1}^{L}
|\sigma_k|^2\right)^{1/2} \left(\sum_{k=1}^{N} |\sigma_k|^2\right)^{-1/2}.
%
\end{equation}
%
This is also called the \textsc{svd} tolerance.  In our application, it relates
to how much \textsc{snr} is lost by discarding $N - L$ of the basis filters
with the lowest singular values.

% Rewritten above.

It is clear that in order to reduce the cost of a low latency search for
compact binaries of unknown mass we need to reduce the number of convolutions
necessary to find the gravitational-wave signal.  Rather than filter the
$i$ templates corresponding to binaries of different component masses it is
possible to find a basis set of fewer filter $u_j$ that can be linearly combined
to reproduce $\rho_i$ to some accuracy.  For the remainder of this section
we will drop the explicit time dependence of the convolution and recognize
that the snr for any point in time is simply the innerproduct between a
template waveform vector ${\bf \hat{h}_i}$ and a vector of the detector data 
${\bf \hat{s}}$, $\rho_i = ({\bf \hat{h}_i}|{\bf \hat{s}})$.  
The goal is to have
\begin{equation}
\label{eq:linear_combination_of_snr}
\rho_i = \sum_{j=1}^{N'} M_{ij} ({\bf \hat{u}_j}|{\bf \hat{s}})
\end{equation}
where the number of inner products is reduced from $N$ to $N'$. In order to
find the basis set ${\bf u_j}$ we use the singular value decomposition of
the matrix 
${H_{ik}} = \{{\bf \hat{h}_1}, {\bf \hat{h}_2}, ...{\bf \hat{h}_N}\}$. Where
$i$ is the index over the template number and $k$ is the index over sample 
points. The singular value decomposition factors a matrix such that
\begin{equation}
\label{eq:svd}
{H_{ik}} = {A_{ij}\,B_{jl}\,U_{lk}}
\end{equation}
The matrix ${B_{jl}}$ is a diagonal matrix of singular values ranked in
order of importance of reconstructing the original matrix ${H_{ik}}$.
However since a search for binary inspirals only needs  waveform accuracy to a
few percent to be successful it is possible to make an approximate 
reconstruction of ${H_{ik}}$ 
\begin{equation}
\label{eq:svdapprox}
{H_{ik}} \approx {A_{ij}\,B'_{jl}\,U_{lk}}
\end{equation}
where the rank of ${B'_{jl}}$ is less than the rank of ${B_{jl}}$.
This reduces the number of rows in ${U_{lk}}$.  Now we have a 
new basis of template waveforms derived from ${U_{lk}} = \{{\bf \hat{u}_1}, {\bf \hat{u}_2}, ...{\bf \hat{u}_{N'}}\}$ and can write 
(\ref{eq:linear_combination_of_snr})
\begin{eqnarray}
\label{eq:svdsnr}
\rho_i &=& \sum_{j=1}^{N'} A_{il}\,B'_{lj} ({\bf \hat{u}_j}|{\bf \hat{s}}) \nonumber \\
       &=& \sum_{j=1}^{N'} M_{ij} ({\bf \hat{u}_j}|{\bf \hat{s}})
\end{eqnarray}

\end{comment}

\editorial{This sounds confrontational to me, but these are good citations.  Can we tone it down?}  This result differs from other work that models gravitational-wave chirp signals in approximate ways~\cite{Chassande-Mottin2006, Candes2008, BuonannoChenVallisneri:2003a} by starting with an exact representation of the desired template family and producing a rigorous approximation with a tunable accuracy.

@@

The composite detection statistic that we employ is a weighted sum of squares, $\Gamma = \sum_{k=1}^L w_k (\rho_k^\perp) ^ 2$, with the particular weights
\begin{equation}
w_k = \frac{\sigma_k^2}{\sigma_k^2 + N / A^2}.
\end{equation}
\editorial{We now have a pretty good description of time slices, the SVD, and conditional reconstruction.  We still need to tie it all together.}Here, $A$ is a desired \textsc{snr} scale that is set by the analyst.  As the authors show in ~\cite{svd-compdetstat}, this choice of weights has a better receiver operating characteristic for a signal of \textsc{snr}~=~$A$ than some other obvious choices, $w_k = \sigma_k^2$ or $w_k = 1$.

% Fold this into SVD discussion.

%\subsection{Detection statistic}
It is not necessary to reconstruct the SNR of all templates in order to 
determine if a signal is present.  By thresholding on the sum of squares
of the orthoganal decomposition described above it is possible to answer
whether or not an inspiral like event is in the data.  Then only 
potentially interesting times can be reconstructed on a sample by sample 
basis.  Following the ideas in~\cite{Anderson:2000yy} we employ a detection
statistic as
\begin{equation}
\label{eq:stat}
\Gamma^2 = \sum_{l} \sum_{j} {B'_{lj}}^2 {({\bf \hat{u}_j}|{\bf \hat{s}})}^2
\end{equation}

\end{comment}

% We aren't using this chi square test.

\subsection{Signal consistency check}
After reconstructing the SNR of an individual template it is possible to 
to check that the components along the SVD basis were consistent with 
expectation by performing a $\chi^2$ fit to the decomposition.  
\begin{equation}
\label{eq:chisq}
\chi^2_i = \sum_j {[({\bf \hat{u}_j}|{\bf \hat{s}}) - \rho_i / M_{ij}]}^2
\end{equation}

\end{comment}

% I think that this part can be swalled into the computational costs discussion above.

\subsection{Comparison of computational costs and latency}
Convolutions are best done with the use of the Fast Fourier Transform (FFT).
However there is an inherent latency in performing convolutions this way. This
section describes the computational cost and latency of FFT convolutions, time 
domain (TD) convolutions and TD convolutions with SVD.

Convolving one filter with data using FFTs involves 
$\mathcal{O}[10 N_s \log_2(N_s)]$
floating point operations where $N_s$ is the number of sample points in
the data being convolved and results in $N_s - N_f$ samples of usable output.  
In comparison time domain convolutions require
$\mathcal{O}[N_s N_f]$ where $N_f$ is the number of sample points in the
filter.  The FFT is faster to compute when $N_f > 10 \log_2(N_s)$.  However
the FFT has an inherent latency of $\mathcal{O}[N_s/f_s]$ seconds, where
$f_s$ is the sample rate when minimally overlapping FFTs are computed. It is
possible to reduce the latency of the FFT by performing vastly overlapping
FFTs.  The cost of such a procedure would be 
$\mathcal{O}[10 N_o N_s \log_2(N_s)]$.  Thus for comparable computational
cost where $N_f \approx 10 N_o \log_2(N_s)$ the latency of the FFT method 
would be $\mathcal{O}[N_s/f_s/N_o]$ seconds.  Depending on the task at hand,
latency requirements and computational costs can dictate what procedure to
use.

\end{comment}

>>>>

In the rest of the section, we will describe a detection strategy that makes
use of some very general properties of inspiral template banks in order to
evaluate a matched filter bank with far lower computational cost than the
conventional \textsc{fir} method and far less latency than the conventional
\textsc{fft} method.

@@

If the template is known to be quasi-bandlimited, then the partial matched
filter for each time slice may be processed at a reduced sample rate.  Compact
binary inspiral waveforms fit this description superbly: they are chirps whose
frequency and amplitude rise according to power laws of time.  They are not
only quasi-bandlimited, but quasi-monochromatic.  By exploiting this property,
we can implement a matched filter for an inspiral waveform using a multirate
filter bank that requires far fewer floating point operations.

@@

Our second innovation exploits the fact that the templates in inspiral template
banks are, by design, highly correlated.  This is due to the fact that points
are chosen in parameter space to be templates such that any point from the
region parameter space we cover has a match to the nearest template greater
than our chosen minimum match. It is possible to greatly reduce the number of
matched filters required to achieve a particular minimum match by designing an
appropriate set of orthonormal {\em basis templates}.  A purely numerical
technique based on the application of the singular value decomposition
(\textsc{svd}) to inspiral waveforms is demonstrated by the authors
in~\cite{Cannon:2010p10398}.

One may regard a bank of $N/2$ discretely-sampled, complex-valued templates of
length $M$ samples, $h^\mathrm W(\theta_i; t_j) = [\mathbf H]_{ij} = H_{ij}$,
as a real matrix of shape $N \times M$, where we have unpacked each
complex-valued template into two real-valued templates.  The singular value
decomposition is an exact factorization that exists for any matrix such that
%
\begin{equation} \label{eq:svd-exact}
%
H_{ij} = [\mathbf {V \Sigma U}]_{ij} = \sum_{k=1}^{N} v_{ik} \sigma_k u_{kj},
%
\end{equation}
%
where $\mathbf V$ and $\mathbf U$ are both unitary matrices and $\mathbf
\Sigma$ is a diagonal matrix.  For our purposes, we associate the rows of
$\mathbf U$ with a minimal set of basis templates, which become the kernels of
basis filters.  These filters give rise to the orthogonal \textsc{snr}s,
$\rho_k^\perp = u_k \star x^\mathsf{W}$.  The rows of $\mathbf{V \Sigma}$
become reconstruction coefficients that map linear combinations of the
orthogonal \textsc{snr}s from the basis filters back onto \textsc{snr}s for the
original templates of interest: $\rho_i = \sum_k v_{ik} \sigma_k \rho_k^\perp$.

In many applications of the \textsc{svd}, including ours, the matrix can be
well approximated by truncating the summation in equation~(\ref{eq:svd-exact})
at $N' \ll N$:
%
\begin{equation}
%
H_{ij}^\prime = \sum_{k=1}^{N'} v_{ik} \sigma_k u_{kj}
%
\end{equation}
%
The cumulative sum of squares of the {\em singular values}, $\sigma_k$,
measures the Frobenius norm of the approximation, such that
%
\begin{equation}
%
\frac{\| \mathbf H^\prime \|}{\| \mathbf H \|} = \left(\sum_{k=1}^{N'}
|\sigma_k|^2\right)^{1/2} \left(\sum_{k=1}^{N} |\sigma_k|^2\right)^{-1/2}.
%
\end{equation}
%
This is also called the \textsc{svd} tolerance.  In our application, it relates
to how much \textsc{snr} is lost by discarding $N - N'$ of the basis filters
with the lowest singular values.

This result differs from other work that models gravitational-wave chirp
signals in approximate ways~\cite{Chassande-Mottin2006, Candes2008,
BuonannoChenVallisneri:2003a} by starting with an exact representation of the
desired template family and producing a rigorous approximation with a tunable
accuracy.

@@

Following the ideas in~\cite{Anderson:2000yy}, we employ is a weighted sum of
squares, $\Gamma = \sum_{k=1}^{N'} w_k (\rho_k^\perp) ^ 2$, with the particular
weights
%
\begin{equation}
%
w_k = \frac{\sigma_k^2}{\sigma_k^2 + N / A^2},
%
\end{equation}
%
where $A^2$ is a desired \textsc{snr} scale that is set by the analyst.  As the
authors show in~\cite{svd-compdetstat}, this choice of weights has a better
receiver operating characteristic for a signal of \textsc{snr}~=~$A$ than some
other obvious choices, $w_k = \sigma_k^2$ or $w_k = 1$.

\subsection{Comparison of computational costs and latency}

Using all of these innovations, we are able to greatly reduce the compuational
cost of searching for these signals. An example of this can be seen in
table~\ref{table:comp_cost} where we compute the computational cost of
filtering 256 templates from a $97\%$ minimal match template bank whitened with
an (FIXME: is this correct?) Advanced \textsc{ligo} \textsc{asd}. It is
interesting to note that by themselves, both the \textsc{svd} and time-slice
innovations \emph{increase} the overall computational cost when combined with
\textsc{fft} filtering. It is only when combined with conditional
reconstruction do these methods yield computational savings. Details of the
computational cost of different operations can be found in \ref{appendix:cost}.

\editorial{It would be good to illustrate the layout of this particular
template bank: masses spanned, time slice layout \dots}
\begin{table}[!h]
\caption{Operation counts per sample for six different detection methods.  The
operation counts for \textsc{lloid} assume a reconstruction duty cycle of 5\%.
Note that the \textsc{fft} method with \textsc{lloid} is almost 10 times faster
than the conventional \textsc{fft} method, despite having substantially lower
latency.}
\label{table:comp_cost}
\begin{center}
\input{flop_budget}
\end{center}
\end{table}
